---
title: "LASSO Logistic Regression for Second-Date Decisions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(20260730)
```

## Research question

Can we predict whether one participant will choose to see another participant
again from racial preferences and other characteristics known before the date?

This is a **binary classification** problem. `decision = 1` means that the
participant said Yes (`dec == 1`), and `decision = 0` means No. The model below
is a LASSO **logistic** regression (`family = "binomial"`), not a linear
regression. A predicted probability of at least 0.5 is classified as Yes.

## Data preparation

The raw file is the source used by `cleaning_data.Rmd`. For this prediction
task, post-date ratings such as `attr`, `sinc`, `intel`, `fun`, `amb`, `shar`,
`like`, `prob`, and `met` are deliberately excluded: they are measured after
the date and would leak information about the outcome. `iid` is also excluded
because it is an identifier, not a characteristic.

```{r packages-and-data}
if (!requireNamespace("glmnet", quietly = TRUE)) {
  stop("This report requires the glmnet package. Install it with install.packages('glmnet').")
}

library(glmnet)

speed_dating_data <- read.csv(
  "data/Speed_Dating_Data.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

model_variables <- c(
  "dec", "gender", "age", "race", "age_o", "race_o", "samerace",
  "imprace", "imprelig", "goal", "date", "go_out", "career_c",
  "sports", "tvsports", "exercise", "dining", "museums", "art",
  "hiking", "gaming", "clubbing", "reading", "tv", "theater",
  "movies", "concerts", "music", "shopping", "yoga", "exphappy"
)

dat <- speed_dating_data[, model_variables]
dat$decision <- as.integer(dat$dec == 1)
dat$dec <- NULL

# Codes represent categories for these variables, not continuous quantities.
categorical_variables <- c(
  "gender", "race", "race_o", "goal", "date", "go_out", "career_c"
)
dat[categorical_variables] <- lapply(dat[categorical_variables], factor)

# Complete-case analysis keeps the preprocessing transparent and prevents
# information from the test set entering imputation or model fitting.
dat <- dat[complete.cases(dat), ]
dat$decision <- factor(dat$decision, levels = c(0, 1))

table(dat$decision)
```

## Train/test split and LASSO fit

The split is stratified by the outcome so that both sets retain the same Yes/No
composition. `cv.glmnet` selects the penalty using 10-fold cross-validation on
the training set. `lambda.1se` is used because it gives the most regularized
model whose cross-validated performance is within one standard error of the
best model, usually producing a more interpretable set of predictors.

```{r split-and-fit}
yes_rows <- which(dat$decision == 1)
no_rows <- which(dat$decision == 0)
train_rows <- c(
  sample(yes_rows, floor(0.80 * length(yes_rows))),
  sample(no_rows, floor(0.80 * length(no_rows)))
)

train <- dat[train_rows, ]
test <- dat[-train_rows, ]

# Build the design matrix once so train and test have identical dummy-variable
# columns even when a rare category is absent from one split.
x_all <- model.matrix(decision ~ ., dat)[, -1, drop = FALSE]
x_train <- x_all[train_rows, , drop = FALSE]
x_test <- x_all[-train_rows, , drop = FALSE]
y_train <- as.integer(as.character(train$decision))
y_test <- as.integer(as.character(test$decision))

cv_lasso <- cv.glmnet(
  x_train, y_train,
  family = "binomial",
  alpha = 1,
  type.measure = "auc",
  nfolds = 10,
  standardize = TRUE
)

lambda_used <- cv_lasso$lambda.1se
lasso_fit <- glmnet(
  x_train, y_train,
  family = "binomial",
  alpha = 1,
  lambda = lambda_used,
  standardize = TRUE
)

probability <- as.numeric(predict(lasso_fit, newx = x_test, type = "response"))
predicted <- as.integer(probability >= 0.5)

plot(cv_lasso, main = "10-fold cross-validation for LASSO logistic regression")
```

The cross-validation graph shows the relationship between the penalty and
cross-validated performance. The dashed vertical lines identify the penalty
chosen by `lambda.min` and the more conservative `lambda.1se` used here.

```{r lasso-path, fig.height=7, fig.width=9}
plot(
  cv_lasso$glmnet.fit,
  xvar = "lambda",
  label = TRUE,
  main = "LASSO coefficient paths"
)
abline(v = log(lambda_used), lty = 2, col = "#B2182B")
```

The coefficient-path graph shows how predictors enter or leave the model as
the penalty changes. The selected model is evaluated at the red vertical line.

## Confusion matrix and classification metrics

```{r classification-metrics}
confusion_matrix <- table(
  Actual = factor(y_test, levels = c(0, 1)),
  Predicted = factor(predicted, levels = c(0, 1))
)
print(confusion_matrix)

tn <- confusion_matrix["0", "0"]
fp <- confusion_matrix["0", "1"]
fn <- confusion_matrix["1", "0"]
tp <- confusion_matrix["1", "1"]

accuracy <- (tp + tn) / sum(confusion_matrix)
precision <- if ((tp + fp) == 0) NA_real_ else tp / (tp + fp)
recall <- if ((tp + fn) == 0) NA_real_ else tp / (tp + fn)
f1_score <- if (is.na(precision) || is.na(recall) || precision + recall == 0) {
  NA_real_
} else {
  2 * precision * recall / (precision + recall)
}

metrics <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
  Value = c(accuracy, precision, recall, f1_score)
)
knitr::kable(metrics, digits = 4)
```

## ROC curve and ROC-AUC

```{r roc}
roc_curve <- function(actual, score) {
  ord <- order(score, decreasing = TRUE)
  actual <- actual[ord]
  tpr <- c(0, cumsum(actual == 1) / sum(actual == 1), 1)
  fpr <- c(0, cumsum(actual == 0) / sum(actual == 0), 1)
  # Remove duplicate points only for a cleaner plot; AUC uses the full path.
  list(fpr = fpr, tpr = tpr)
}

roc <- roc_curve(y_test, probability)
roc_auc <- sum(diff(roc$fpr) * (head(roc$tpr, -1) + tail(roc$tpr, -1)) / 2)

plot(roc$fpr, roc$tpr, type = "s", lwd = 2, col = "#2166AC",
     xlab = "False positive rate", ylab = "True positive rate",
     main = sprintf("ROC curve (AUC = %.3f)", roc_auc))
abline(0, 1, lty = 2, col = "grey50")
roc_auc
```

## Precision-recall curve and PR-AUC

```{r pr}
pr_curve <- function(actual, score) {
  ord <- order(score, decreasing = TRUE)
  actual <- actual[ord]
  tp <- cumsum(actual == 1)
  fp <- cumsum(actual == 0)
  precision <- tp / (tp + fp)
  recall <- tp / sum(actual == 1)
  list(precision = c(1, precision), recall = c(0, recall))
}

pr <- pr_curve(y_test, probability)
pr_auc <- sum(diff(pr$recall) * (head(pr$precision, -1) + tail(pr$precision, -1)) / 2)

plot(pr$recall, pr$precision, type = "s", lwd = 2, col = "#B2182B",
     xlab = "Recall", ylab = "Precision",
     main = sprintf("Precision-recall curve (AUC = %.3f)", pr_auc))
abline(h = mean(y_test == 1), lty = 2, col = "grey50")
pr_auc
```

## Calibration plot

ROC-AUC and PR-AUC evaluate ranking performance. A calibration plot checks
whether predicted probabilities are numerically reasonable: among observations
assigned probabilities near 0.2, for example, roughly 20% should be Yes.

```{r calibration}
calibration_data <- data.frame(actual = y_test, probability = probability)
calibration_data$bin <- cut(
  calibration_data$probability,
  breaks = seq(0, 1, by = 0.1),
  include.lowest = TRUE
)
calibration_summary <- aggregate(
  cbind(actual, probability) ~ bin,
  data = calibration_data,
  FUN = mean
)

plot(
  calibration_summary$probability,
  calibration_summary$actual,
  type = "b",
  pch = 19,
  lwd = 2,
  col = "#4D9221",
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Mean predicted probability",
  ylab = "Observed proportion saying Yes",
  main = "Calibration plot"
)
abline(0, 1, lty = 2, col = "grey50")
```

## Variables selected by LASSO

The coefficients below are on the log-odds scale. Because glmnet standardizes
predictors while fitting, coefficients are interpreted per one standardized
unit of a predictor. For factor variables, each coefficient compares that
level with the omitted reference level. The odds ratio is `exp(coefficient)`.

```{r coefficients}
coefficient_matrix <- as.matrix(coef(lasso_fit))
coefficient_table <- data.frame(
  Variable = rownames(coefficient_matrix),
  Coefficient = as.numeric(coefficient_matrix[, 1]),
  row.names = NULL
)
coefficient_table <- subset(coefficient_table, Variable != "(Intercept)" & Coefficient != 0)
coefficient_table$Odds_Ratio <- exp(coefficient_table$Coefficient)
coefficient_table$Direction <- ifelse(
  coefficient_table$Coefficient > 0,
  "Increases probability of Yes",
  "Decreases probability of Yes"
)
coefficient_table <- coefficient_table[order(coefficient_table$Coefficient, decreasing = TRUE), ]

knitr::kable(coefficient_table, digits = 4)
```

```{r selected-coefficients, fig.height=7, fig.width=9}
if (nrow(coefficient_table) > 0) {
  old_par <- par(mar = c(5, 11, 4, 2) + 0.1)
  coefficient_table$Variable <- factor(
    coefficient_table$Variable,
    levels = coefficient_table$Variable[order(coefficient_table$Coefficient)]
  )
  barplot(
    coefficient_table$Coefficient,
    names.arg = coefficient_table$Variable,
    horiz = TRUE,
    las = 1,
    col = ifelse(coefficient_table$Coefficient > 0, "#2166AC", "#B2182B"),
    main = "Nonzero LASSO coefficients",
    xlab = "Coefficient on the log-odds scale",
    cex.names = 0.75
  )
  abline(v = 0, lty = 2)
  par(old_par)
}
```

Blue bars indicate variables associated with higher odds of a Yes decision;
red bars indicate lower odds, conditional on the other predictors. For factor
variables, the coefficient is relative to the reference category.

Positive coefficients (odds ratios greater than 1) increase the predicted
probability of saying Yes, holding the other model terms constant. Negative
coefficients (odds ratios below 1) decrease it. The table is the report of the
selected variables and their coefficient values; variables absent from the
table were shrunk exactly to zero.

```{r model-summary}
cat("Lambda.1se:", lambda_used, "\n")
cat("Selected non-intercept variables:", nrow(coefficient_table), "of", ncol(x_train), "\n")
```

## Why coefficients were shrunk to zero

LASSO adds an L1 penalty to the logistic-regression loss. As the penalty grows,
the model can set weak or redundant coefficients exactly to zero. In this
dataset, several pre-date variables are correlated (for example, overlapping
activity preferences), and race-related indicators can carry overlapping
information. LASSO therefore keeps a sparse subset rather than assigning a
separate nonzero coefficient to every correlated input. A zero coefficient does
not prove that a variable has no relationship with dating decisions; it means
that, at the cross-validated penalty and conditional on the other predictors,
the variable did not add enough out-of-sample predictive value to be retained.
The `lambda.1se` choice intentionally favors this simpler interpretation.
