---
title: "Logistic Regression for Second-Date Decisions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(20260730)
```

## Research question

Can we predict whether one participant will choose to see another participant
again from the cleaned demographic, racial-context, and date-evaluation fields?

This is a binary classification problem. `decision = 1` means Yes
(`dec == 1`) and `decision = 0` means No. The model is an ordinary
ordinary binomial logistic regression, not a continuous regression. A predicted
probability of at least 0.50 is classified as Yes.

## Data preparation

The model uses exactly the cleaned fields supplied for the project:
`gender`, `age`, `race`, `age_o`, `race_o`, `samerace`, `attr`,
`sinc`, `intel`, `fun`, `amb`, and `shar`. `iid` is an identifier
and is excluded. `dec` is converted into the outcome and is not a predictor.

```{r data}
speed_dating_data <- read.csv(
  "data/Speed_Dating_Data.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

model_variables <- c(
  "dec", "gender", "age", "race", "age_o", "race_o", "samerace",
  "attr", "sinc", "intel", "fun", "amb", "shar"
)

dat <- speed_dating_data[, model_variables]
dat$decision <- as.integer(dat$dec == 1)
dat$dec <- NULL

categorical_variables <- c("gender", "race", "race_o")
dat[categorical_variables] <- lapply(dat[categorical_variables], factor)

dat <- dat[complete.cases(dat), ]
dat$decision <- factor(dat$decision, levels = c(0, 1))
table(dat$decision)
```

## Train/test split and logistic-regression fit

The split is stratified by the Yes/No outcome. The model is fit on 80% of the
data, and the remaining 20% is held out for final evaluation.

```{r fit-logistic-regression}
yes_rows <- which(dat$decision == 1)
no_rows <- which(dat$decision == 0)
train_rows <- c(
  sample(yes_rows, floor(0.80 * length(yes_rows))),
  sample(no_rows, floor(0.80 * length(no_rows)))
)

train <- dat[train_rows, , drop = FALSE]
test <- dat[-train_rows, , drop = FALSE]
y_test <- as.integer(as.character(test$decision))

logistic_fit <- glm(
  decision ~ .,
  data = train,
  family = binomial()
)

probability_yes <- as.numeric(predict(logistic_fit, newdata = test, type = "response"))
classification_threshold <- 0.50
predicted <- as.integer(probability_yes >= classification_threshold)
```

## Confusion matrix and classification metrics

```{r metrics}
confusion_matrix <- table(
  Actual = factor(y_test, levels = c(0, 1), labels = c("No", "Yes")),
  Predicted = factor(predicted, levels = c(0, 1), labels = c("No", "Yes"))
)
confusion_matrix

tp <- confusion_matrix["Yes", "Yes"]
tn <- confusion_matrix["No", "No"]
fp <- confusion_matrix["No", "Yes"]
fn <- confusion_matrix["Yes", "No"]

safe_divide <- function(numerator, denominator) {
  if (denominator == 0) NA_real_ else numerator / denominator
}

accuracy <- safe_divide(tp + tn, sum(confusion_matrix))
precision <- safe_divide(tp, tp + fp)
recall <- safe_divide(tp, tp + fn)
f1_score <- safe_divide(2 * precision * recall, precision + recall)
```

## ROC curve and ROC-AUC

```{r roc}
roc_curve <- function(actual, score) {
  ordering <- order(score, decreasing = TRUE)
  ordered_actual <- actual[ordering]
  data.frame(
    false_positive_rate = c(0, cumsum(ordered_actual == 0) / sum(ordered_actual == 0)),
    true_positive_rate = c(0, cumsum(ordered_actual == 1) / sum(ordered_actual == 1))
  )
}

roc <- roc_curve(y_test, probability_yes)
roc_auc <- sum(diff(roc$false_positive_rate) *
  (head(roc$true_positive_rate, -1) + tail(roc$true_positive_rate, -1)) / 2)

plot(
  roc$false_positive_rate, roc$true_positive_rate,
  type = "s", lwd = 2, col = "#2166AC",
  xlab = "False positive rate", ylab = "True positive rate",
  main = sprintf("ROC curve (ROC-AUC = %.3f)", roc_auc)
)
abline(0, 1, lty = 2, col = "grey50")
legend(
  "bottomright",
  legend = c(sprintf("Logistic regression: ROC-AUC = %.3f", roc_auc),
             "No-discrimination reference"),
  col = c("#2166AC", "grey50"), lty = c(1, 2), lwd = c(2, 1), bty = "n"
)
```

## Precision-recall curve, PRC-AUC, and average precision

```{r pr}
pr_curve <- function(actual, score) {
  ordering <- order(score, decreasing = TRUE)
  ordered_actual <- actual[ordering]
  true_positive <- cumsum(ordered_actual == 1)
  data.frame(
    recall = c(0, true_positive / sum(ordered_actual == 1)),
    precision = c(1, true_positive / seq_along(ordered_actual))
  )
}

pr <- pr_curve(y_test, probability_yes)
prc_auc <- sum(diff(pr$recall) *
  (head(pr$precision, -1) + tail(pr$precision, -1)) / 2)
average_precision <- sum(diff(pr$recall) * tail(pr$precision, -1))

plot(
  pr$recall, pr$precision,
  type = "s", lwd = 2, col = "#B2182B",
  xlab = "Recall", ylab = "Precision",
  main = sprintf("Precision-recall curve (PRC-AUC = %.3f)", prc_auc)
)
abline(h = mean(y_test), lty = 2, col = "grey50")
legend(
  "bottomleft",
  legend = c(
    sprintf("Logistic regression: PRC-AUC = %.3f", prc_auc),
    sprintf("Average precision = %.3f", average_precision),
    sprintf("No-skill average Yes rate = %.3f", mean(y_test))
  ),
  col = c("#B2182B", "#B2182B", "grey50"),
  lty = c(1, 1, 2), lwd = c(2, 2, 1), bty = "n"
)
```

```{r complete-metrics}
metrics <- data.frame(
  Metric = c(
    "Accuracy", "Precision", "Recall", "F1 Score",
    "ROC-AUC", "PRC-AUC", "Average precision"
  ),
  Value = c(
    accuracy, precision, recall, f1_score,
    roc_auc, prc_auc, average_precision
  )
)
knitr::kable(metrics, digits = 4)
```

## Calibration plot

```{r calibration}
bins <- cut(probability_yes, breaks = seq(0, 1, by = 0.10), include.lowest = TRUE)
calibration <- data.frame(
  mean_prediction = as.numeric(tapply(probability_yes, bins, mean)),
  observed_yes_rate = as.numeric(tapply(y_test, bins, mean))
)
calibration <- calibration[complete.cases(calibration), , drop = FALSE]

plot(
  calibration$mean_prediction, calibration$observed_yes_rate,
  type = "b", pch = 19, lwd = 2, col = "#4D9221",
  xlim = c(0, 1), ylim = c(0, 1),
  xlab = "Mean predicted probability", ylab = "Observed Yes rate",
  main = "Calibration plot"
)
abline(0, 1, lty = 2, col = "grey50")
```

## Logistic-regression coefficients

The coefficients are on the log-odds scale. Positive coefficients are
associated with higher odds of a Yes decision, holding all other fields
constant. The odds ratio is `exp(coefficient)`.

```{r coefficients}
coefficient_table <- data.frame(
  Variable = names(coef(logistic_fit)),
  Coefficient = as.numeric(coef(logistic_fit)),
  row.names = NULL
)
coefficient_table <- subset(coefficient_table, Variable != "(Intercept)")
coefficient_table$Odds_Ratio <- exp(coefficient_table$Coefficient)
coefficient_table$Direction <- ifelse(
  coefficient_table$Coefficient > 0,
  "Higher odds of Yes",
  "Lower odds of Yes"
)
coefficient_table <- coefficient_table[order(coefficient_table$Coefficient, decreasing = TRUE), ]

knitr::kable(coefficient_table, digits = 4)
```

```{r coefficient-plot, fig.height=7, fig.width=9}
coefficient_plot_data <- coefficient_table[order(coefficient_table$Coefficient), ]
old_par <- par(mar = c(5, 11, 4, 2) + 0.1)
barplot(
  coefficient_plot_data$Coefficient,
  names.arg = coefficient_plot_data$Variable,
  horiz = TRUE,
  las = 1,
  col = ifelse(coefficient_plot_data$Coefficient > 0, "#2166AC", "#B2182B"),
  main = "Logistic-regression coefficients",
  xlab = "Coefficient on the log-odds scale",
  cex.names = 0.75
)
abline(v = 0, lty = 2)
par(old_par)
```
