---
title: "Feedforward Neural Network for Second-Date Decisions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(20260730)
```

## Research question

Can we predict whether one participant chooses to see another participant again
from tracked racial preferences and other characteristics known before the
date? This is a **binary classification** problem. The model predicts a
probability internally, but the final prediction is always a categorical
`Yes` or `No` decision.

The outcome is defined as follows:

```r
decision = 1  # Yes: the participant wants to see the partner again
decision = 0  # No
```

The neural network uses one hidden layer, so it is a feedforward neural network
and not a regression model. A probability of at least `0.50` is classified as
`Yes`.

## Data preparation

The model uses the cleaned fields supplied for this project: the participant's
and partner's demographics/racial context (`gender`, `age`, `race`, `age_o`,
`race_o`, and `samerace`) plus the participant's ratings of the date (`attr`,
`sinc`, `intel`, `fun`, `amb`, and `shar`). `iid` is excluded because it is an
identifier, and `dec` is converted into the binary outcome rather than used as
a predictor. No interests, goals, dating habits, career, or other variables are
used.

```{r packages-and-data}
if (!requireNamespace("nnet", quietly = TRUE)) {
  stop("This report requires the nnet package. Install it with install.packages('nnet').")
}

library(nnet)

speed_dating_data <- read.csv(
  "data/Speed_Dating_Data.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

model_variables <- c(
  "dec", "gender", "age", "race", "age_o", "race_o", "samerace",
  "attr", "sinc", "intel", "fun", "amb", "shar"
)

dat <- speed_dating_data[, model_variables]
dat$decision <- as.integer(dat$dec == 1)
dat$dec <- NULL

categorical_variables <- c(
  "gender", "race", "race_o"
)
dat[categorical_variables] <- lapply(dat[categorical_variables], factor)

dat <- dat[complete.cases(dat), ]
dat$decision <- factor(dat$decision, levels = c(0, 1), labels = c("No", "Yes"))

table(dat$decision)
```

## Stratified train/test split

The split is stratified so that both sets contain Yes and No decisions in
approximately the same proportions. The test set is held out until the final
evaluation.

```{r split-and-design-matrix}
yes_rows <- which(dat$decision == "Yes")
no_rows <- which(dat$decision == "No")

train_rows <- c(
  sample(yes_rows, floor(0.80 * length(yes_rows))),
  sample(no_rows, floor(0.80 * length(no_rows)))
)

train <- dat[train_rows, , drop = FALSE]
test <- dat[-train_rows, , drop = FALSE]

x_all <- model.matrix(decision ~ ., dat)[, -1, drop = FALSE]
x_train <- x_all[train_rows, , drop = FALSE]
x_test <- x_all[-train_rows, , drop = FALSE]
y_train <- as.integer(train$decision == "Yes")
y_test <- as.integer(test$decision == "Yes")

train_means <- colMeans(x_train)
train_sds <- apply(x_train, 2, sd)
train_sds[is.na(train_sds) | train_sds == 0] <- 1

x_train_scaled <- sweep(sweep(x_train, 2, train_means, "-"), 2, train_sds, "/")
x_test_scaled <- sweep(sweep(x_test, 2, train_means, "-"), 2, train_sds, "/")
```

## Feedforward neural network

The network has one hidden layer with eight units. `entropy = TRUE` fits a
logistic binary output, so `predict(..., type = "raw")` returns the estimated
probability of `Yes`. Weight decay regularizes the network and helps reduce
overfitting.

```{r fit-neural-network}
nn_fit <- nnet(
  x = x_train_scaled,
  y = y_train,
  size = 8,
  decay = 0.001,
  maxit = 500,
  entropy = TRUE,
  linout = FALSE,
  trace = FALSE,
  MaxNWts = 10000
)

probability_yes <- as.numeric(predict(nn_fit, x_test_scaled, type = "raw"))

classification_threshold <- 0.50
predicted_decision <- factor(
  ifelse(probability_yes >= classification_threshold, "Yes", "No"),
  levels = c("No", "Yes")
)
actual_decision <- factor(test$decision, levels = c("No", "Yes"))

confusion_matrix <- table(
  Actual = actual_decision,
  Predicted = predicted_decision
)
confusion_matrix
```

The model's final output is categorical:

```{r categorical-predictions}
predictions <- data.frame(
  probability_yes = probability_yes,
  predicted_decision = predicted_decision,
  actual_decision = actual_decision
)
head(predictions)
```

## Test-set evaluation

```{r evaluation}
tp <- sum(y_test == 1 & predicted_decision == "Yes")
tn <- sum(y_test == 0 & predicted_decision == "No")
fp <- sum(y_test == 0 & predicted_decision == "Yes")
fn <- sum(y_test == 1 & predicted_decision == "No")

safe_divide <- function(numerator, denominator) {
  if (denominator == 0) NA_real_ else numerator / denominator
}

auc <- function(actual, scores) {
  positive <- actual == 1
  n_positive <- sum(positive)
  n_negative <- sum(!positive)
  if (n_positive == 0 || n_negative == 0) return(NA_real_)
  ranks <- rank(scores, ties.method = "average")
  (sum(ranks[positive]) - n_positive * (n_positive + 1) / 2) /
    (n_positive * n_negative)
}

accuracy <- safe_divide(tp + tn, length(y_test))
sensitivity <- safe_divide(tp, tp + fn)
specificity <- safe_divide(tn, tn + fp)
balanced_accuracy <- mean(c(sensitivity, specificity), na.rm = TRUE)

pr_ordering <- order(probability_yes, decreasing = TRUE)
pr_actual <- y_test[pr_ordering]
pr_true_positive <- cumsum(pr_actual == 1)
pr_recall <- c(0, pr_true_positive / sum(pr_actual == 1))
pr_precision <- c(1, pr_true_positive / seq_along(pr_actual))
prc_auc <- sum(diff(pr_recall) *
  (head(pr_precision, -1) + tail(pr_precision, -1)) / 2)
average_precision <- sum(diff(pr_recall) * tail(pr_precision, -1))

metrics <- data.frame(
  threshold = classification_threshold,
  accuracy = accuracy,
  sensitivity_yes = sensitivity,
  specificity_no = specificity,
  balanced_accuracy = balanced_accuracy,
  roc_auc = auc(y_test, probability_yes),
  prc_auc = prc_auc,
  average_precision = average_precision
)
metrics
```

## Important classification graphs

The following graphs summarize the outcome balance, the model's categorical
predictions, discrimination, calibration, and which encoded predictors most
affect held-out accuracy. PNG copies are saved in `graphs/neural_network/`.

```{r create-classification-graphs}
graph_dir <- file.path("graphs", "neural_network")
dir.create(graph_dir, recursive = TRUE, showWarnings = FALSE)

save_png <- function(filename, plot_function, width = 1600, height = 1000) {
  png(file.path(graph_dir, filename), width = width, height = height, res = 180)
  on.exit(dev.off(), add = TRUE)
  plot_function()
}

plot_class_balance <- function() {
  counts <- table(dat$decision)
  barplot(
    counts,
    col = c("#4C78A8", "#F58518"),
    ylab = "Number of decisions",
    main = "Class balance in the modeling data"
  )
}

plot_confusion_matrix <- function() {
  plot.new()
  plot.window(xlim = c(0, 2), ylim = c(0, 2))
  title("Neural network confusion matrix (test set)")
  for (row in seq_len(2)) {
    for (column in seq_len(2)) {
      x_left <- column - 1
      y_bottom <- 2 - row
      fill <- if (row == column) "#B7D8B7" else "#F2C2C2"
      rect(x_left, y_bottom, x_left + 1, y_bottom + 1, col = fill, border = "white")
      text(x_left + 0.5, y_bottom + 0.5, confusion_matrix[row, column], cex = 1.6)
    }
  }
  axis(1, at = c(0.5, 1.5), labels = colnames(confusion_matrix), tick = FALSE)
  axis(2, at = c(1.5, 0.5), labels = rownames(confusion_matrix), tick = FALSE, las = 1)
  mtext("Predicted decision", side = 1, line = 2.4)
  mtext("Actual decision", side = 2, line = 2.4)
  box()
}

roc_curve <- function(actual, scores) {
  ordering <- order(scores, decreasing = TRUE)
  ordered_actual <- actual[ordering]
  positive_total <- sum(ordered_actual == 1)
  negative_total <- sum(ordered_actual == 0)
  data.frame(
    false_positive_rate = c(0, cumsum(ordered_actual == 0) / negative_total),
    true_positive_rate = c(0, cumsum(ordered_actual == 1) / positive_total)
  )
}

pr_curve <- function(actual, scores) {
  ordering <- order(scores, decreasing = TRUE)
  ordered_actual <- actual[ordering]
  true_positive <- cumsum(ordered_actual == 1)
  data.frame(
    recall = c(0, true_positive / sum(ordered_actual == 1)),
    precision = c(1, true_positive / seq_along(ordered_actual))
  )
}

roc_values <- roc_curve(y_test, probability_yes)
pr_values <- pr_curve(y_test, probability_yes)

plot_roc <- function() {
  plot(
    roc_values$false_positive_rate, roc_values$true_positive_rate,
    type = "l", lwd = 3, col = "#4C78A8",
    xlab = "False positive rate", ylab = "True positive rate",
    main = sprintf("ROC curve (ROC-AUC = %.3f)", metrics$roc_auc),
    xlim = c(0, 1), ylim = c(0, 1)
  )
  abline(0, 1, lty = 2, col = "gray50")
  legend(
    "bottomright",
    legend = c(sprintf("Model: ROC-AUC = %.3f", metrics$roc_auc),
               "No-discrimination reference"),
    col = c("#4C78A8", "gray50"), lty = c(1, 2), lwd = c(3, 1), bty = "n"
  )
}

plot_precision_recall <- function() {
  plot(
    pr_values$recall, pr_values$precision,
    type = "l", lwd = 3, col = "#F58518",
    xlab = "Recall (sensitivity for Yes)", ylab = "Precision",
    main = sprintf("Precision-recall curve (PRC-AUC = %.3f)", prc_auc),
    xlim = c(0, 1), ylim = c(0, 1)
  )
  abline(h = mean(y_test), lty = 2, col = "gray50")
  legend(
    "bottomleft",
    legend = c(
      sprintf("Model: PRC-AUC = %.3f; average precision = %.3f", prc_auc, average_precision),
      sprintf("No-skill average Yes rate = %.3f", mean(y_test))
    ),
    col = c("#F58518", "gray50"), lty = c(1, 2), lwd = c(3, 1), bty = "n"
  )
}

plot_probability_distribution <- function() {
  no_scores <- probability_yes[y_test == 0]
  yes_scores <- probability_yes[y_test == 1]
  breaks <- seq(0, 1, by = 0.05)
  no_histogram <- hist(no_scores, breaks = breaks, plot = FALSE)
  yes_histogram <- hist(yes_scores, breaks = breaks, plot = FALSE)
  ymax <- max(c(no_histogram$counts, yes_histogram$counts))
  plot(
    no_histogram, col = rgb(0.30, 0.47, 0.66, 0.55), border = "white",
    xlim = c(0, 1), ylim = c(0, ymax),
    xlab = "Predicted probability of Yes", ylab = "Test-set count",
    main = "Predicted-probability distribution by actual decision"
  )
  plot(yes_histogram, col = rgb(0.96, 0.52, 0.09, 0.55), border = "white", add = TRUE)
  abline(v = classification_threshold, lty = 2, lwd = 2)
  legend(
    "topright", legend = c("Actual No", "Actual Yes", "0.50 threshold"),
    fill = c(rgb(0.30, 0.47, 0.66, 0.55), rgb(0.96, 0.52, 0.09, 0.55), NA),
    border = c("white", "white", NA), lty = c(NA, NA, 2), bty = "n"
  )
}

plot_calibration <- function() {
  bins <- cut(probability_yes, breaks = seq(0, 1, by = 0.10), include.lowest = TRUE)
  calibration <- data.frame(
    predicted_probability = as.numeric(tapply(probability_yes, bins, mean)),
    observed_yes_rate = as.numeric(tapply(y_test, bins, mean))
  )
  calibration <- calibration[complete.cases(calibration), , drop = FALSE]
  plot(
    calibration$predicted_probability, calibration$observed_yes_rate,
    pch = 19, cex = 1.2, col = "#54A24B",
    xlab = "Mean predicted probability", ylab = "Observed Yes rate",
    main = "Calibration plot (test set)", xlim = c(0, 1), ylim = c(0, 1)
  )
  lines(calibration$predicted_probability, calibration$observed_yes_rate,
        col = "#54A24B", lwd = 2)
  abline(0, 1, lty = 2, col = "gray50")
}

set.seed(20260730)
baseline_accuracy <- mean(as.integer(predicted_decision == "Yes") == y_test)
permutation_importance <- vapply(seq_len(ncol(x_test_scaled)), function(column) {
  permuted_test <- x_test_scaled
  permuted_test[, column] <- sample(permuted_test[, column])
  permuted_probability <- as.numeric(predict(nn_fit, permuted_test, type = "raw"))
  permuted_prediction <- as.integer(permuted_probability >= classification_threshold)
  baseline_accuracy - mean(permuted_prediction == y_test)
}, numeric(1))
names(permutation_importance) <- colnames(x_test_scaled)

importance <- sort(permutation_importance, decreasing = TRUE)

plot_permutation_importance <- function() {
  barplot(
    rev(importance), names.arg = rev(names(importance)),
    horiz = TRUE, las = 1, col = "#72B7B2",
    xlab = "Decrease in test accuracy after permutation",
    main = "Permutation importance of encoded predictors"
  )
  abline(v = 0, lty = 2, col = "gray50")
}

save_png("01_class_balance.png", plot_class_balance)
save_png("02_confusion_matrix.png", plot_confusion_matrix)
save_png("03_roc_curve.png", plot_roc)
save_png("04_precision_recall_curve.png", plot_precision_recall)
save_png("05_probability_distribution.png", plot_probability_distribution)
save_png("06_calibration_plot.png", plot_calibration)
save_png("07_permutation_importance.png", plot_permutation_importance)

graph_files <- file.path(
  graph_dir,
  c(
    "01_class_balance.png", "02_confusion_matrix.png", "03_roc_curve.png",
    "04_precision_recall_curve.png", "05_probability_distribution.png",
    "06_calibration_plot.png", "07_permutation_importance.png"
  )
)
```

```{r display-classification-graphs, fig.show='hold', out.width='95%'}
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::include_graphics(graph_files)
} else {
  graph_files
}
```

## Conclusion

This analysis uses a feedforward neural network for a binary categorical
outcome. The network estimates `P(decision = Yes)`, and the fixed threshold of
0.50 converts that probability into the required categorical `Yes`/`No`
prediction. The reported accuracy, sensitivity, specificity, balanced
accuracy, confusion matrix, ROC-AUC, PRC-AUC, and average precision are all
classification results evaluated on held-out observations.
